Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-02-08 pkg/disttree/inst/draft/draft.Rnw
edited reference for GLM
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1995 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e3035a23815155ee68617f65719cc6eebbf7ef6e
small corrections in the intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1997 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 51e9aa7d193d49f77f5be8a452a8756bd1c0cf84

In regression analysis various models have been developed to describe the
relationship between a response variable and a set of covariates. The classical
model is the linear model (LM) where the conditional mean of the response
is modeled through a linear function of the covariates (see the left panel of
Figure~\ref{Devel_parmod} for a schematic illustration). Over the last decades
this has been extended in various directions including:
\begin{itemize}
  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1983})
    encompassing an additional nonlinear link function for the conditional mean.
  \item Generalized additive models (GAMs, \citealp{Hastie+Tibshirani:1986})
    allowing for smooth nonlinear effects in the covariates
    (Figure~\ref{Devel_parmod}, middle).
  \item Generalized additive models for location, scale, and shape
    (GAMLSS, \citealt{stasinopoulos2005}) adopting a probabilistic modeling
    approach. In GAMLSS, each parameter of a statistical distribution can depend
    on an additive predictor of the covariates comprising linear and/or
    smooth nonlinear terms (Figure~\ref{Devel_parmod}, right).
\end{itemize}
Thus, the above-mentioned models provide a broad toolbox for capturing different
aspects of the response (mean only vs.\ full distribution) and different types
of dependencies on the covariates (linear vs.\ nonlinear additive terms).

@@

While in many applications conditional mean regression models had been receiving
the most attention, there has been a paradigm shift over the last decade towards
distributional regression models. An important reason for this is that in many
fields forecasts of the mean are not the only (or not even the main) concern but
that there is an increasing interest in probabilistic forecasts. Quantities of
interest typically include exceedence probabilities for certain thresholds of the
response or quantiles of the response distribution. As an example, consider
weather forecasting where there is less interest in the mean amount of
precipitation on the next day. Instead, the probability of rain vs.\ no rain
is typically more relevant or, in some situations, a prediction interval of
expected precipitation (say from the expected 10\% to 90\% quantile). Similar
considerations apply for other meteorological quantities and hence attention
in the weather forecasting literature has been shifting from classical linear
models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
non-homogenous Gaussian regression (NGR) of \cite{gneiting2005calibrated}.
The NGR typically describes the mean of some meteorological response variable through
the average of the corresponding quantity from an ensemble of physically-based
numerical weather predictions (NWPs). Similarly, the variance of the response
is captured through the variance of the ensemble of NWP outputs. Thus, the NGR
considers both the mean as well as the uncertainty of the NWP model outputs
to obtain probabilistic forecasts calibrated to a particular site.
%% late: introduce MOS and EMOS? more details about precipitation forecasting

@@

The before discussed models with the response depending on additive linear or smooth nonlinear terms
of the covariates are powerful tools as long as only smooth effects are to be considered.
A different approach that can also deal with non-additive effects
and possibly unknown interactions of covariates is to apply tree-based methods. 
A regression tree (\citealt{breiman1984classification})
works on the idea of separating the data into more homogeneous subgroups based on
covariates. However, the resulting step function as illustrated in the left panel of
Figure \ref{Devel_treeforest} can often be too rough and imposes abrupt changes on the model/data even in
case of smooth transitions. Random forest models such as Breiman and
Cutler's randomForest (\citealt{breiman2001random}) encounter this problem/disadvantage by combining an
ensemble of trees such that steps are smoothed out in case of smooth effects while abrupt changes
are still captured by its tree-based structure (see the top middel panel of Figure \ref{Devel_treeforest}).
While classic regression trees and random forests only model the mean of the response
we are now going to combine the idea of these tree based-methods with the progress
made in the field of (flexible) parametric modelling. In particular, the development from mean regression
models to distributional models will be embeded in the framework or regression trees and
random forests providing distributional trees and forests. This development is illustrated
in the bottom panel of Figure~\ref{Devel_treeforest} for distributional trees and in the top right panel
for distributional forests.

In many modeling situations, especially when the type of effects are not known in advance,
it can be challenging to specify a well fitting model. The novel methodology of distributional 
trees and forests provides a framework that covers a wide range of effects including linear and nonlinear
additive and non-additive effects as well as interactions of covariates. Moreover, these models are
particularly advantageous in case of many covariates with unknown effects or
interactions as the variable selection is done automatically within the model
fitting process. Therefore, no additional effort or knowledge for the
specification of the model is required in advance.

For these reasons, distributional trees and forests provide
valuable features which can contribute to improving models for probabilistic
forecasting. For example in precipitation/weather forecasting (for complex terrain) 
in mountainous regions various influencing factors can change rapidly and their effects are 
difficult to be represented by a smooth parametric model. (Additionally, a wide grid over
this complex terrain can lead to abrupt changes within just one cell.)

In the application discussed in this paper precipitation forecasts are made for a mountainous region
in the Alps covering mainly Tyrol, Austria, and adjacent areas. Predictions are based on data from 
85 observation stations and 28 years (1985--2012). The total amount of precipitation within 24 hours
is modeled with the assumption of the underlying distribution being a censored normal distribution 
with censoring point 0. As covariates a large number of meteorlogical quantities are provided by 
ensembles of numerical weather prediction models. 
Next to distributional other methods commonly used in this research area are applied and their results are compared. 
In a first setting only one observation station is taken into consideration. The models are trained on data from 
24 years and probabilistic predictions of the consecutive 4 years are validated. A brief outlook on the results of 
this setting is given by Figure~\ref{axams_onestation}. For the 24th of July of 2009, 2010, 2011 and 2012 
the density functions of the distributions estimated by a distributional forest are plotted together with the 
corresponding observations. This illustrates that appropriate probabilistic forecasts 
are provided by this model as they are close to the observed values. Beside the expected values the predictions 
provided by distributional models offer a wide range of information as they specify the whole distribution and 
therefore enable further methods of inference and analysis. In section \ref{ProbabilisticForecasting} this example 
will be further discussed and it will be shown that the predictions made by the distributional forest are 
reasonably well fitting, not only for the mean parameter of the distribution but also for its second parameter, 
the scale parameter, and therefore the whole distribution.

@@

@
\caption{\label{boxplot_crps_all}CRPS skills score for each station (grey lines) and aggregated over all stations (boxplots). Station Axams is highlighted in greend and the horizontal orange line pertains to the reference model EMOS. The models are learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
\end{center}
\end{figure}

@@

\item After Bonferroni adjusting the p-vlaues check whether any of the resulting p-values is beneath the selected significance level. If so, choose the partitioning variable $Z_{j^\ast}$ with the lowest p-value to the scores corresponding to the distribution parameters relevant for the split.

>>>>

In regression analysis various models have been developed to describe the
relationship between a response variable and a set of covariates. The classical
model is the linear model (LM) where the conditional mean of the response
is modeled through a linear function of the covariates (see the left panel of
Figure~\ref{Devel_parmod} for a schematic illustration). Over the last decades
this has been extended in various directions including:
\begin{itemize}
  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1989})
    encompassing an additional nonlinear link function for the conditional mean.
  \item Generalized additive models (GAMs, \citealp{Hastie+Tibshirani:1986})
    allowing for smooth nonlinear effects in the covariates
    (Figure~\ref{Devel_parmod}, middle).
  \item Generalized additive models for location, scale, and shape
    (GAMLSS, \citealt{stasinopoulos2005}) adopting a probabilistic modeling
    approach. In GAMLSS, each parameter of a statistical distribution can depend
    on an additive predictor of the covariates comprising linear and/or
    smooth nonlinear terms (Figure~\ref{Devel_parmod}, right).
\end{itemize}
Thus, the above-mentioned models provide a broad toolbox for capturing different
aspects of the response (mean only vs.\ full distribution) and different types
of dependencies on the covariates (linear vs.\ nonlinear additive terms).

@@

While in many applications conditional mean regression models had been receiving
the most attention, there has been a paradigm shift over the last decade towards
distributional regression models. An important reason for this is that in many
fields forecasts of the mean are not the only (or not even the main) concern but
that there is an increasing interest in probabilistic forecasts. Quantities of
interest typically include exceedence probabilities for certain thresholds of the
response or quantiles of the response distribution. As an example, consider
weather forecasting where there is less interest in the mean amount of
precipitation on the next day. Instead, the probability of rain vs.\ no rain
is typically more relevant or, in some situations, a prediction interval of
expected precipitation (say from the expected 10\% to 90\% quantile). Similar
considerations apply for other meteorological quantities and hence attention
in the weather forecasting literature has been shifting from classical linear
models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
non-homogeneous Gaussian regression (NGR) of \cite{gneiting2005calibrated}.
The NGR typically describes the mean of some meteorological response variable through
the average of the corresponding quantity from an ensemble of physically-based
numerical weather predictions (NWPs). Similarly, the variance of the response
is captured through the variance of the ensemble of NWP outputs. Thus, the NGR
considers both the mean as well as the uncertainty of the NWP model outputs
to obtain probabilistic forecasts calibrated to a particular site.
%% late: introduce MOS and EMOS? more details about precipitation forecasting

@@

The before discussed models with the response depending on additive linear or smooth nonlinear terms
of the covariates are powerful tools as long as only smooth effects are to be considered.
A different approach that can also deal with non-additive effects
and possibly unknown interactions of covariates is to apply tree-based methods. 
A regression tree (\citealt{breiman1984classification})
works on the idea of separating the data into more homogeneous subgroups based on
covariates. However, the resulting step function as illustrated in the left panel of
Figure \ref{Devel_treeforest} can often be too rough and impose abrupt changes on the model even in
case of smooth transitions. Random forest models such as Breiman and
Cutler's randomForest (\citealt{breiman2001random}) encounter this problem by combining an
ensemble of trees such that steps are smoothed out in case of smooth effects while abrupt changes
are still captured by its tree-based structure (see the top middle panel of Figure \ref{Devel_treeforest}).

While classic regression trees and random forests only model the mean of the response
we are now going to combine the idea of these tree based-methods with the progress
made in the field of flexible parametric modelling. In particular, the development from mean regression
models to distributional models will be embedded in the framework of regression trees and
random forests providing distributional trees and forests. This development is illustrated
in the bottom panel of Figure~\ref{Devel_treeforest} for distributional trees and in the top right panel
for distributional forests.

In many modeling situations, especially when the types of effects are not known in advance,
it can be challenging to specify a well fitting model. The novel methodology of distributional 
trees and forests provides a framework that covers a wide range of effects including linear and nonlinear
additive and non-additive effects as well as interactions of covariates. Moreover, these models are
particularly advantageous in case of many covariates with unknown effects or
interactions as the variable selection is done automatically within the model
fitting process. Therefore, no additional effort or knowledge for the
specification of the model is required in advance.

For these reasons, distributional trees and forests provide
valuable features which can contribute to improving models for probabilistic
forecasting. For example weather forecasting in mountainous regions various influencing factors 
can change rapidly and their effects are  difficult to be represented by a smooth parametric model. 
Additionally, large spacing of the grid applied in numerical weather prediction models over 
complex terrain can lead to abrupt changes within just one cell.

In the application discussed in this paper precipitation forecasts are made for a mountainous region
in the Alps covering mainly Tyrol in Austria and adjacent areas. Predictions are based on data from 
85 observation stations and 28 years (1985--2012). The total amount of precipitation within 24 hours
is modeled with the assumption of the underlying distribution being a censored normal distribution 
with censoring point 0. As covariates a large number of meteorological quantities are provided by 
ensembles of numerical weather prediction models. 
Next to distributional forests other methods commonly used in this research area are applied and their 
results are compared. In a first setting only one observation station is taken into consideration.
The models are trained on data from 24 years and probabilistic predictions of the consecutive 4 years 
are validated. A brief outlook on the results of this setting is given by Figure~\ref{axams_onestation}. 
For the 24th of July of 2009, 2010, 2011 and 2012 the density functions of the distributions estimated 
by a distributional forest are plotted together with the corresponding observations. This illustrates 
that appropriate probabilistic forecasts are provided by this model as they are close to the observed 
values. Beside the expected values the predictions made by distributional models offer a wide range 
of information as they specify the whole distribution and therefore enable further methods of inference 
and analysis. In section \ref{ProbabilisticForecasting} this application will be further discussed and it 
will be shown that the predictions provided by the distributional forest are reasonably well fitting, not 
only for the mean parameter of the distribution but also for its second parameter, the scale parameter, 
and therefore the whole distribution.

@@

@
\caption{\label{boxplot_crps_all}CRPS skills score for each station (grey lines) and aggregated over all stations (boxplots). Station Axams is highlighted in green and the horizontal orange line pertains to the reference model EMOS. The models are learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
\end{center}
\end{figure}

@@

\item After Bonferroni adjusting the p-values check whether any of the resulting p-values is beneath the selected significance level. If so, choose the partitioning variable $Z_{j^\ast}$ with the lowest p-value to the scores corresponding to the distribution parameters relevant for the split.
